---
title: "STAT 425 Final Project - Statistical Analysis"
author: "Zoya Malik, Manreet Kaur Jhattu, Lachlan Findlay"
date: "April 2026"
output: html_notebook
---

```{r setup, include=FALSE}
library(ggplot2)
library(mosaic)
library(car)
options(scipen = 999)
```

---

# 1. Load and Inspect the Data

```{r}
# read in the combined dataset
typing.df <- read.csv("data/typing_combined.csv")

# convert treatment and block to factors
typing.df$Music  <- as.factor(typing.df$Music)
typing.df$Tester <- as.factor(typing.df$Tester)

str(typing.df)
```

```{r}
# quick look at the data
head(typing.df, 10)
```

---

# 2. Summary Statistics

Summary statistics for WPM broken down by Music condition (the treatment):

```{r}
favstats(WPM ~ Music, data = typing.df)
```

Summary statistics for WPM broken down by Tester (the block):

```{r}
favstats(WPM ~ Tester, data = typing.df)
```

---

# 3. Data Visualization

## Boxplots by Music Condition

```{r}
ggplot(data = typing.df, aes(x = Music, y = WPM)) +
  geom_boxplot(col = "red", fill = "light blue") +
  xlab("Auditory Condition") +
  ylab("Words Per Minute (WPM)") +
  ggtitle("Boxplots:", subtitle = "Typing Speed Across Auditory Conditions") +
  coord_flip()
```

## Boxplots by Tester (Block)

```{r}
ggplot(data = typing.df, aes(x = Tester, y = WPM)) +
  geom_boxplot(col = "red", fill = "light blue") +
  xlab("Tester") +
  ylab("Words Per Minute (WPM)") +
  ggtitle("Boxplots:", subtitle = "Typing Speed Across Testers") +
  coord_flip()
```

---

# 4. The Mathematical Model

We are carrying out a **Complete Randomized Block Design (CRBD)**. The mathematical model is:

$$Y_{ij} = \mu + \tau_i + \beta_j + \varepsilon_{ij}$$

where:

- $Y_{ij}$ is the mean WPM for the $i$th auditory condition and $j$th tester
- $\mu$ is the overall mean WPM
- $\tau_i$ is the effect of the $i$th auditory condition ($i = 1, 2, 3, 4$ for Lo-Fi, Silence, Upbeat, White Noise), with $\sum_{i=1}^{4} \tau_i = 0$
- $\beta_j$ is the effect of the $j$th tester/block ($j = 1, 2, 3$ for Manreet, Lachlan, Zoya), with $\sum_{j=1}^{3} \beta_j = 0$
- $\varepsilon_{ij} \overset{iid}{\sim} N(0, \sigma^2)$ are the random error terms

Both the auditory condition (treatment) and the tester (block) are fixed factors, as all four music conditions were deliberately chosen and the three testers are the specific individuals in the study.

---

# 5. Checking Model Conditions

Before running the ANOVA, we check the three conditions of the CRBD model normality, equal variance, and additivity.

## Fit the model and extract residuals

```{r}
# fit the CRBD model: treatment + block
typing.aov <- aov(WPM ~ Music + Tester, data = typing.df)

# extract residuals and fitted values
e_terms     <- residuals(typing.aov)
fitted_vals <- fitted(typing.aov)

# add to data frame for plotting
typing.diag.df <- cbind(typing.df, e_terms, fitted_vals)
```

## Condition 1: Normality of Residuals

```{r}
# normal probability plot of residuals
ggplot(data = typing.diag.df, aes(sample = e_terms)) +
  stat_qq(size = 2, col = "blue") +
  stat_qqline(col = "red") +
  ggtitle("Normal Probability Plot of the Residuals")
```

```{r}
# Shapiro-Wilk test for normality
shapiro.test(e_terms)
```

## Condition 2: Equal Variance

```{r}
# residual plot - residuals vs fitted values
ggplot(data = typing.diag.df, aes(x = fitted_vals, y = e_terms)) +
  geom_point(size = 2, col = "blue") +
  xlab("Fitted Values") +
  ylab("Residuals") +
  geom_hline(yintercept = 0, linetype = "dashed", col = "red") +
  ggtitle("Residual Plot")
```

```{r}
# Levene's test for equal variance across Music conditions
leveneTest(WPM ~ Music, data = typing.df)
```

## Condition 3: Additivity (No Interaction Between Treatment and Block)

```{r}
# interaction plot - Music on x-axis, Tester as trace
interaction.plot(x.factor   = typing.df$Music,
                 trace.factor = typing.df$Tester,
                 response    = typing.df$WPM,
                 type        = "b",
                 pch         = c(1, 2, 4),
                 col         = c("red", "blue", "darkgreen"),
                 lwd         = 2,
                 xlab        = "Auditory Condition",
                 ylab        = "Mean WPM",
                 trace.label = "Tester",
                 main        = "Interaction Plot: Auditory Condition x Tester")
```

---

# 6. CRBD Analysis of Variance

## Statistical Hypotheses

**For the Auditory Condition (Treatment) Effect:**

$$H_0: \tau_1 = \tau_2 = \tau_3 = \tau_4 = 0 \quad \text{(no auditory condition effect on WPM)}$$
$$H_A: \text{At least one } \tau_i \neq 0$$

**For the Tester (Block) Effect:**

$$H_0: \beta_1 = \beta_2 = \beta_3 = 0 \quad \text{(no tester effect on WPM)}$$
$$H_A: \text{At least one } \beta_j \neq 0$$

## ANOVA Table

```{r}
# CRBD ANOVA: treatment + block
summary(typing.aov)
```

## Extract F-statistics and P-values

```{r}
# extract results for the music condition effect
anova.results <- summary(typing.aov)[[1]]

F_music <- anova.results["Music", "F value"]
p_music <- anova.results["Music", "Pr(>F)"]
cat("F_Obs for Music Condition:", round(F_music, 4), "\n")
cat("P-value for Music Condition:", round(p_music, 4), "\n")
```

```{r}
# extract results for the tester (block) effect
F_tester <- anova.results["Tester", "F value"]
p_tester <- anova.results["Tester", "Pr(>F)"]
cat("F_Obs for Tester:", round(F_tester, 4), "\n")
cat("P-value for Tester:", round(p_tester, 4), "\n")
```

---

# 7. Multiple Comparisons - Tukey HSD

If the auditory condition effect is significant, we use the Tukey HSD method to determine which specific conditions differ from one another.

```{r}
# Tukey HSD on the Music (treatment) factor
tukey.music <- TukeyHSD(typing.aov, "Music", conf.level = 0.95)
tukey.music
```

```{r}
# plot the Tukey HSD results
plot(tukey.music, las = 1)
```

```{r}
# group means by Music condition, sorted descending
sort(favstats(WPM ~ Music, data = typing.df)$mean, decreasing = TRUE)
```

---

# 8. Summary of Results

```{r}
# overall mean WPM by condition and tester
tapply(typing.df$WPM, list(typing.df$Music, typing.df$Tester), mean)
```

```{r}
# marginal means by Music condition
tapply(typing.df$WPM, typing.df$Music, mean)
```

```{r}
# marginal means by Tester
tapply(typing.df$WPM, typing.df$Tester, mean)
```
